A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-22 tex/codatadp.tex
more

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\AST$ is a completion of $\Val$ for normal model $\M$, the following holds:
\begin{enumerate}
\item[\rm 1.] $\Val^\AST \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm 2.]
$\Val^\AST \ec{x} = \Val^\AST \ec{y} \interp{ t }$ for
all subterms $t$ of $\Val^\AST \ec{y}$ with $\muvar(t) = {\Varec{x}}$. 
\item[\rm 3.]
$\Val^\AST \ec{x} \vsim \Val^\AST \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{report}
\begin{proof}
To show 1, we first show that $\Val^\AST$ contains no free variables.
Assume by contradiction that $\Val^\AST$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$ to ${\Varec{y}}$.
By construction, $\Val^\AST \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\AST$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\Val^\AST \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\AST \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

We first show that 2 and 3 hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_{\negvthinspace j}( \vec t )$ and $\const{C}_{\vthinspace\jPrime}( \vec u )$ in $\ec{z}$
are such that $j = \jPrime$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply to $\Ec$.
Thus, 
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$ 
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_{\negvthinspace j}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_{\negvthinspace j}( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $t$ that are bound in $\Val \ec{y}$ are interpreted in 
the expansion of $\Val \ec{y} \interp{ t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. \rem{more detail}
Thus, we have
$\Val \ec{y} \interp{ t } = \Val \ec{x} \interp{ \Val \ec{x} } = \Val \ec{x}$, and thus 2 holds for $\Val$.
Statement 3 holds for $\Val$ since \rn{Unique} does not apply.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \Val^\AST \ec{ u_1 }, \ldots, \Val^\AST \ec{ u_n } )}$, call this term $t_u$.
%By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\AST$, we have $\Val^\AST \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\Val^\AST \ec{t}\interp{w_i} = \Val^\AST \ec{u_i}$ for $i = 1, \ldots, n$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by our construction of $\Val^\AST$ we have $\Val^\AST \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$ 
where for each $i = 1, \ldots, n$, 
$\muvar(w_i) = {\Varec{u_i}}$ and thus by Lemma~\ref{lem:model-completion}.2, 
$\Val^\AST \ec{t}\interp{w_i} = \Val^\AST \ec{u_i}$.
For each $i = 1, \ldots, n$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $t_u\interp{u_\iPrime} \vsim \Val^\AST \ec{u_i}$.
Thus $t_u\interp{u_\iPrime} \vsim \Val^\AST \ec{t}\interp{ w_i }$. % for each $i = 1, \ldots, n$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \Val^\AST \ec{t}$, and
we have $\M( t ) \vsim \Val^\AST \ec{t}$.

If $t$ is a selector term of the form $\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\const C_{\vthinspace\jPrime}\bigl( \const s^1_{\vthinspace\jPrime}( u ), \ldots, \const s^{n}_{\vthinspace\jPrime}( u ) \bigr)$ for some $\jPrime$.
%By Lemma~\ref{lem:model-completion}.2 and due to our construction of $\Val^\AST$, we have $\Val^\AST \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\vthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\Val^\AST \ec{u} \interp{ w_i } = \Val^\AST \ec{ \const s^i_{\vthinspace\jPrime}( u )}$ for $i = 1, \ldots, n$. 
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by our construction of $\Val^\AST$ we have $\Val^\AST \ec{u}$ is of the form
$\MU {\Varec{u}}.\; \const C_{\vthinspace\jPrime}( w_1, \ldots, w_n )$,
where for each $i = 1, \ldots, n$,
$\muvar(w_i) = {\Varec{\const s^i_{\vthinspace\jPrime}( u )}}$ and thus by Lemma~\ref{lem:model-completion}.2, 
$\Val^\AST \ec{u} \interp{ w_i } = \Val^\AST \ec{ \const s^i_{\vthinspace\jPrime}( u )}$. 
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\Val^\AST \ec{u} \interp{w_k}$, which is equal to $\Val^\AST \ec{ \const s^k_{\negvthinspace j}( u )}$,
which is $\Val^\AST \ec{t}$.
If $j \neq \jPrime$, we are free to interpret $\const s^k_{\negvthinspace j}( \M( u ) )$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\Val^\AST \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}.3,
pairs of wrongly applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\M( t_1 ) \vsim \Val^\AST \ec{t_1}$ and $\M( t_2 ) \vsim \Val^\AST \ec{t_2}$,
by Lemma~\ref{lem:model-completion}.3, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.

@@

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_{\negvthinspace j}( t )$ is such that $\const{d}_{\negvthinspace j}$
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^n_{\negvthinspace j}( t ) \bigr)$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
if $\const s^k_{\negvthinspace j}( t )$ is a term in $\tEc$ and $t$ contains the constructor
$\const{C}_{\negvthinspace j}( \vec u )$, we immediately infer $\const s^k_{\negvthinspace j}( t ) \teq u_k$, whereas the calculus we presented would apply \rn{Split} and \rn{Inject} before inferring this equality.
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a preprocessing step.
In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_{\negvthinspace j}( \vec t ) \teq \const{C}_{\vthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}( \const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$,
and replace all occurrences of discriminators $\const{d}_{\negvthinspace j}( \const{C}_{\vthinspace\jPrime}( \vec t ) )$ by $\top$ if $j = \jPrime$ and $\bot$ otherwise.

>>>>

%\newcommand\PART[1]{.#1}
\newcommand\PART[1]{(#1)}

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\AST$ is a completion of $\Val$ for normal model $\M$, the following properties hold:
\begin{enumerate}
\item[\rm (1)] $\Val^\AST \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item[\rm (2)]
$\Val^\AST \ec{x} = \Val^\AST \ec{y} \interp{ t }$ for
all subterms $t$ of $\Val^\AST \ec{y}$ with $\muvar(t) = {\Varec{x}}$. 
\item[\rm (3)]
$\Val^\AST \ec{x} \vsim \Val^\AST \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{report}
\begin{proof}
To show 1, we first show that $\Val^\AST$ contains no free variables.
Assume by contradiction that $\Val^\AST$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be $1$.
Since \rn{Single} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{y}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_{\Varec{y}}(\Val^\AST)$ to ${\Varec{y}}$.
By construction, $\Val^\AST \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\AST$ applies substitutions of the form
$\{ {\Varec{y}} \mapsto t \}$, where $t$ is acyclic when the type of ${\Varec{y}}$ is not a codatatype.
Thus, $\Val^\AST \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\AST \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

We first show that 2 and 3 hold initially for $\Val$.
For all equivalence classes $\ec{z}$,
each pair of constructor terms $\const{C}_{\negvthinspace j}( \vec t )$ and $\const{C}_{\negvthinspace\jPrime}( \vec u )$ in $\ec{z}$
are such that $j = \jPrime$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Inject} does not apply to $\Ec$.
Thus, 
$\Val$ was constructed by applying a sequence of substitutions
where all substitutions for variables ${\Varec{z}}$ 
were uniquely of the form $\{ {\Varec{z}} \mapsto \const{C}_{\negvthinspace j}( {\Varec{t_1}}, \ldots, {\Varec{t_n}} ) \}$
when $\ec{z}$ contains a constructor $\const{C}_{\negvthinspace j}( t_1, \ldots, t_n )$.
Say $\Val \ec{y}$ has a subterm $t$ where $\muvar(t) = {\Varec{x}}$.
Both $\Val \ec{x}$ and the subterm $t$ of $\Val \ec{y}$ were
constructed by applying a sequence of substitutions of the form mentioned above to ${\Varec{x}}$.
Moreover, free variables ${\Varec{z}}$ in $t$ that are bound in $\Val \ec{y}$ are interpreted in 
the expansion of $\Val \ec{y} \interp{ t }$ as a term
constructed by a sequence of substitutions of the form mentioned above to ${\Varec{z}}$. \rem{more detail}
Thus, we have
$\Val \ec{y} \interp{ t } = \Val \ec{x} \interp{ \Val \ec{x} } = \Val \ec{x}$, and thus 2 holds for $\Val$.
Statement 3 holds for $\Val$ since \rn{Unique} does not apply.

@@

If $t$ is a constructor term of the form $\const{C}( u_1, \ldots, u_n )$,
then $\M( t )$ is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \M( u_1 ), \ldots, \M( u_n ) ) }$ for some fresh $x$,
which by the induction hypothesis is $\alpha$-equivalent with $\nf{\MU x.\; \const{C}( \Val^\AST \ec{ u_1 }, \ldots, \Val^\AST \ec{ u_n } )}$, call this term $t_u$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\Val^\AST$, we have $\Val^\AST \ec{t}$ is a term
%of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$
%where $\Val^\AST \ec{t}\interp{w_i} = \Val^\AST \ec{u_i}$ for $i = 1, \ldots, n$.
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by our construction of $\Val^\AST$ we have $\Val^\AST \ec{t}$ is a term
of the form $\MU {\Varec{t}}.\; \const{C}( w_1, \ldots, w_n )$ 
where for each $i = 1, \ldots, n$, 
$\muvar(w_i) = {\Varec{u_i}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2}, 
$\Val^\AST \ec{t}\interp{w_i} = \Val^\AST \ec{u_i}$.
For each $i = 1, \ldots, n$, let $u_\iPrime$ be the $i$\vvthinspace th argument of $t_u$.
By Lemma~\ref{lem:mu-norm-arg}, $t_u\interp{u_\iPrime} \vsim \Val^\AST \ec{u_i}$.
Thus $t_u\interp{u_\iPrime} \vsim \Val^\AST \ec{t}\interp{ w_i }$. % for each $i = 1, \ldots, n$.
%By Lemma~\ref{lem:mu-cong}, we have $t_u \vsim t_w$.
Thus, $\M( t ) \vsim t_u \vsim \Val^\AST \ec{t}$, and
we have $\M( t ) \vsim \Val^\AST \ec{t}$.

If $t$ is a selector term of the form $\const s^k_{\negvthinspace j}( u )$,
since \rn{Split} does not apply to $\Fc$,
$\ec{u}$ must contain a term of the form $\const C_{\negvthinspace\jPrime}\bigl( \const s^1_{\negvthinspace\jPrime}( u ), \ldots, \const s^{n}_{\negvthinspace\jPrime}( u ) \bigr)$ for some $\jPrime$.
%By Lemma~\ref{lem:model-completion}\PART{2} and due to our construction of $\Val^\AST$, we have $\Val^\AST \ec{u}$ is of the form
%$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
%where $\Val^\AST \ec{u} \interp{ w_i } = \Val^\AST \ec{ \const s^i_{\negvthinspace\jPrime}( u )}$ for $i = 1, \ldots, n$. 
Since \rn{Inject} and \rn{Clash} do not apply to $\Fc$, 
by construction $\Val^\AST$ must be of the form
$\MU {\Varec{u}}.\; \const C_{\negvthinspace\jPrime}( w_1, \ldots, w_n )$,
where for each $i = 1, \ldots, n$,
$\muvar(w_i) = \smash{\Varec{\vphantom{I}\smash{\const s^i_{\negvthinspace\jPrime}( u )}}}$ and thus by Lemma~\ref{lem:model-completion}\PART{2}, 
$\Val^\AST \ec{u} \interp{ w_i } = \Val^\AST \ec{ \smash{\const s^i_{\negvthinspace\jPrime}( u )}}$. 
If $j = \jPrime$, then $\M( t )$ is $\alpha$-equivalent with $\Val^\AST \ec{u} \interp{w_k}$, which is equal to $\Val^\AST \ec{ \smash{ \const s^k_{\negvthinspace j}( u )}}$,
which is $\Val^\AST \ec{t}$.
If $j \neq \jPrime$, we are free to interpret $\smash{\const s^k_{\negvthinspace j}( \M( u ) )}$ as the value in $\M( \tau )$ that is $\alpha$-equivalent with $\Val^\AST \ec{ t }$.
Since \rn{Cong} does not apply and due to Lemma~\ref{lem:model-completion}\PART{3},
pairs of wrongly applied selectors are assigned the same value if and only if they reside in the same equivalence class,
and thus this definition is well-defined.

We now show that all disequalities in $\Fc$ are satisfied by $\M$.
Say $t_1 \tneq t_2 \in \Fc$.
Since \rn{Conflict} does not apply, $t_1 \teq t_2 \not\in \Fc$ and thus $\ec{t_1}$ and $\ec{t_2}$ are distinct.
Since $\M( t_1 ) \vsim \Val^\AST \ec{t_1}$ and $\M( t_2 ) \vsim \Val^\AST \ec{t_2}$,
by Lemma~\ref{lem:model-completion}\PART{3}, $\M( t_1 ) \neq \M( t_2 )$, and thus $\M \models t_1 \tneq t_2$.

@@

Discriminators are treated as predicate symbols, so a discriminator $\const{d}_{\negvthinspace j}( t )$ is such that $\const{d}_{\negvthinspace j}$
is a predicate symbol, instead of reducing this constraint to $t \teq \const{C}_{\negvthinspace j}\bigl( \const s^1_{\negvthinspace j}( t ), \ldots, \const s^n_{\negvthinspace j}( t ) \bigr)$.
We have found that this leads to better performance since this reduction introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several additional rules (see~\cite{} for more details), which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
if $\const s^k_{\negvthinspace j}( t )$ is a term in $\tEc$ and $t$ contains the constructor
$\const{C}_{\negvthinspace j}( \vec u )$, we immediately infer $\const s^k_{\negvthinspace j}( t ) \teq u_k$, whereas the calculus we presented would apply \rn{Split} and \rn{Inject} before inferring this equality.
For the sake of reducing the number of unique constraints considered by the calculus, we compute a normal form for constraints as a preprocessing step.
In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term ordering,
replace $\const{C}_{\negvthinspace j}( \vec t ) \teq \const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}( \const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$,
and replace all occurrences of discriminators $\const{d}_{\negvthinspace j}( \const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ if $j = \jPrime$ and $\bot$ otherwise.
